---
title: "Practical Machine Learning Course Project"
author: "Gopal Ojha"
date: "July 24, 2015"
output: html_document
---

## Prepare the datasets
Read the training data into a data table.
```{r}
library(data.table)
setInternet2(TRUE)
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
D <- fread(url)
```
Read the testing data into a data table.
```{r}
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
DTest <- fread(url)
```
Belt, arm, dumbbell, and forearm variables that do not have any missing values in the test dataset will be predictor candidates.
```{r}
isAnyMissing <- sapply(DTest, function (x) any(is.na(x) | x == ""))
isPredictor <- !isAnyMissing & grepl("belt|[^(fore)]arm|dumbbell|forearm", names(isAnyMissing))
predCandidates <- names(isAnyMissing)[isPredictor]
predCandidates
```
Subset the primary dataset to include only the predictor candidates and the outcome variable, classe.
```{r}
varToInclude <- c("classe", predCandidates)
D <- D[, varToInclude, with=FALSE]
dim(D)
names(D)
```
Make classe into a factor.
```{r}
D <- D[, classe := factor(D[, classe])]
D[, .N, classe]
```
Split the dataset into a 60% training and 40% probing dataset.
```{r}
library(caret)
seed <- as.numeric(as.Date("2015-07-24"))
set.seed(seed)
inTrain <- createDataPartition(D$classe, p=0.6)
DTrain <- D[inTrain[[1]]]
DProbe <- D[-inTrain[[1]]]
```
Preprocess the prediction variables by centering and scaling.
```{r}
X <- DTrain[, predCandidates, with=FALSE]
preProc <- preProcess(X)
preProc
XCS <- predict(preProc, X)
DTrainCS <- data.table(data.frame(classe = DTrain[, classe], XCS))
```
Apply the centering and scaling to the probing dataset.
```{r}
X <- DProbe[, predCandidates, with=FALSE]
XCS <- predict(preProc, X)
DProbeCS <- data.table(data.frame(classe = DProbe[, classe], XCS))
```
Check for near zero variance.
```{r}
nzv <- nearZeroVar(DTrainCS, saveMetrics=TRUE)
if (any(nzv$nzv)) nzv else message("No variables with near zero variance")
```
## Train a prediction model
Using random forest. The error will be estimated using the 40% probing sample.

Set up the parallel clusters.
```{r}
library(parallel)
library(doParallel)
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)
```
Set the control parameters.
```{r}
ctrl <- trainControl(classProbs=TRUE,
                     savePredictions=TRUE,
                     allowParallel=TRUE)
```
Fit model over the tuning parameters.
```{r}
method <- "rf"
system.time(trainingModel <- train(classe ~ ., data=DTrainCS, method=method))
```
Stop the clusters.
```{r}
stopCluster(cl)
```
## Evaluate the model on the training dataset
```{r}
trainingModel
hat <- predict(trainingModel, DTrainCS)
confusionMatrix(hat, DTrain[, classe])
```
## Evaluate the model on the probing dataset
```{r}
hat <- predict(trainingModel, DProbeCS)
confusionMatrix(hat, DProbeCS[, classe])
```
## Display the final model
```{r}
varImp(trainingModel)
trainingModel$finalModel
```
The estimated error rate is less than 1%.

Save training model object for later.
```{r}
save(trainingModel, file="trainingModel.RData")
```
## Predict on the test data
Load the training model.
```{r}
load(file="trainingModel.RData", verbose=TRUE)
```

## Get predictions and evaluate.
```{r}
DTestCS <- predict(preProc, DTest[, predCandidates, with=FALSE])
hat <- predict(trainingModel, DTestCS)
DTest <- cbind(hat , DTest)
subset(DTest, select=names(DTest)[grep("belt|[^(fore)]arm|dumbbell|forearm", names(DTest), invert=TRUE)])
```
